---
title: "Metabarcoding_analysis"
author: "Menuka Bhandari"
format: html
self-contained: true
editor_options: 
  chunk_output_type: console
---

# Setup

Concept of reverse complement- explain it by using example
Forward strand:  5′- ATGCCA -3′
Reverse strand:  3′- TACGGT -5′
```{r}
a <- "ATGCCA"
dna <-DNAString(a)
dna
reverse(dna)
reverseComplement(dna)
```

## Load all the packages needed for analysis
```{r}
# Load packages
library(dada2)
library(phyloseq)
library(tidyverse)
```

## Read all files
```{r}
# Save the number of cores in a variable
n_cores <- 24

# Dir with input FASTQ files (post-cutadapt trimmed reads):
indir <- "results/cutadapt"

# FASTA file with training data - a database your sequences will be compared to (SILVA, UNITE, etc):
# (Check for an up-to-date version at <https://benjjneb.github.io/dada2/training.html>)
tax_key <- "data/ref/silva_nr99_v138.1_train_set.fa.gz"  
```

## Define output directories
```{r}
# Define the output dirs
outdir <- "results/dada"                     
filter_dir <- "results/dada/filtered_fastq"

# Create directories if they are not already created:
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
dir.create(filter_dir, showWarnings = FALSE, recursive = TRUE)

# Create vectors with FASTQ file names
fastqs_raw_F <- sort(list.files(indir, pattern = "_R1_001.fastq.gz", full.names = TRUE))
fastqs_raw_R <- sort(list.files(indir, pattern = "_R2_001.fastq.gz", full.names = TRUE))
```

# Filtering (remove poor quality reads) and quality trimming (poor quality bases)
- truncLen : Truncate reads after truncLen bases and discard reads shorter than this
- trimLeft : trim the first 10 nucleotides of each read 
- maxEE : trim reads based on the maximum numbers of Expected Errors (EE) given the quality scores of the reads’ bases
trunQ - trim read after quality bases 2 or less than 2

```{r eval= FALSE}
fastqs_filt_F <- file.path(filter_dir, fastqs_raw_F)
fastqs_filt_R <- file.path(filter_dir, fastqs_raw_R)

# Filter the FASTQ files
filter_results <-
  filterAndTrim(fastqs_raw_F, fastqs_filt_F,
                fastqs_raw_R, fastqs_filt_R,
                truncLen = c(250,210),
                trimLeft = 10,
                maxN = 0,
                maxEE = c(2,2),
                truncQ = 2,
                rm.phix = FALSE,
                multithread = n_cores,
                compress = FALSE, verbose = TRUE) 

```

# Dereplication and error training 
Dereplication : condense the data by collapsing together all reads that encode the same sequence. Learn error : learns error from each sample.
```{r}
# Dereplicate the FASTQ files
fastqs_derep_F <- derepFastq(fastqs_filt_F, verbose = FALSE)
fastqs_derep_R <- derepFastq(fastqs_filt_R, verbose = FALSE)
# Error training
err_F <- learnErrors(fastqs_derep_F, multithread = n_cores, verbose = TRUE)
err_R <- learnErrors(fastqs_derep_R, multithread = n_cores, verbose = TRUE)
# Plot the error profiles
plotErrors(err_F, nominalQ = TRUE)
plotErrors(err_R, nominalQ = TRUE)
```

# Infer ASVs 
Infer ASVs from unique sequences. Explore pool options to see the different available options.
```{r}
# Infer ASVs
dada_Fs <- dada(fastqs_derep_F, err = err_F, pool = FALSE, multithread = n_cores)
dada_Rs <- dada(fastqs_derep_R, err = err_R, pool = FALSE, multithread = n_cores)
dada_Fs[[1]]
```

# Merge read pairs :  merge the forward and reverse reads together to obtain the 
full denoised sequences. If the forward and reverse reads overlap by at least 
**12 bases**, merging will happen by default.

```{r}
# Merge forward and reverse reads
mergers <- mergePairs(dada_Fs, fastqs_derep_F,
                      dada_Rs, fastqs_derep_R,
                      verbose = TRUE)
```


# Remove chimera
```{r}
# Save the dereplicated objects
saveRDS(fastqs_derep_F, file = file.path(outdir, "fastqs_derep_F.rds"))
saveRDS(fastqs_derep_R, file = file.path(outdir, "fastqs_derep_R.rds"))

rm(fastqs_derep_F, fastqs_derep_R) # Remove objects from environment
```

# Generate summary table
```{r}
# Create the sequence table
seqtab_all <- makeSequenceTable(mergers)
# Check the dimensions of the sequence table
dim(seqtab_all)
# Check the ASV sequence lengths
table(nchar(getSequences(seqtab_all)))
```

# Remove chimera
```{r}
# Identify and remove chimeras
seqtab <- removeBimeraDenovo(seqtab_all,
                             method = "consensus",
                             multithread = n_cores,
                             verbose = TRUE)
# Proportion of retained sequences:
sum(seqtab) / sum(seqtab_all)
# Save the sequence table
saveRDS(seqtab, file = file.path(outdir, "seqtab.rds"))
```

# Generate a summary table
```{r}
# Get the numbers of unique reads at each step
getN <- function(x) sum(getUniques(x))

denoised_F <- sapply(dada_Fs, getN)
denoised_R <- sapply(dada_Rs, getN)
merged <- sapply(mergers, getN)
# Create a summary table
nreads_summary <- data.frame(filter_results,
                             denoised_F,
                             denoised_R,
                             merged,
                             nonchim = rowSums(seqtab),
                             row.names = c("504A0", "504A14", "504B0", "504B14"))
colnames(nreads_summary)[1:2] <- c("input", "filtered")
```

```{r}
# Write the summary table to file
write.table(nreads_summary,
            file = file.path(outdir, "nreads_summary.txt"),
            sep = "\t", quote = FALSE, row.names = TRUE)
```



**Considerations for your own data: This is a great place to do a last sanity check**. 
Outside of filtering, there should no step in which a majority of reads are lost. If a majority of reads failed to merge, you may need to revisit the truncLen parameter used in the filtering step and make sure that the truncated reads span your amplicon. If a majority of reads were removed as chimeric, you may need to revisit the removal of primers, as the ambiguous nucleotides in unremoved primers interfere with chimera identification.**



























